SmartCheck: Automatic and Efficient Counterexample Reduction and Generalization

### 2014-06-23 paper/paper.tex
paper: final version edits.

%% \conferenceinfo{ICFP'12,} {September 9--15, 2012, Copenhagen, Denmark.}
%% \CopyrightYear{2012}
%% \copyrightdata{978-1-4503-1054-3/12/09}

@@

%% \category{D.2.4}{Software/Program Verification}{Reliability}

@@

%% \keywords
%% embedded domain-specific language, compiler, verification

@@

In addition, shrinking is one side of the coin when it comes to making
counterexamples more understandable: the other side is extrapolation from
individual counterexamples to a class of counterexamples.  This leap of
abstraction is often implicitly made by the programmer in determining the reason
why counterexamples fail the property.  For example, the following is a
relatively small counterexample returned when QuickChecking a property in (a
bug-injected version of) \xmonad, a popular X11 window manager written in
Haskell~\cite{xmonad}:
%
\medskip%
\begin{sloppypar}
%\footnotesize
\noindent%
\ttp{StackSet \{current = Screen \{workspace = Workspace \{tag = NonNegative
      \{getNonNegative = 0\}, layout = -1, stack = Just (Stack \{focus = 'S', up =
        "", down = ""\})\}, screen = 1, screenDetail = 1\}, visible = [Screen
    \{workspace = Workspace \{tag = NonNegative \{getNonNegative = 2\}, layout = -1,
        stack = Nothing\}, screen = 2, screenDetail = -1\},Screen \{workspace =
      Workspace \{tag = NonNegative \{getNonNegative = 3\}, layout = -1, stack =
        Nothing\}, screen = 0, screenDetail = -1\}], hidden = [Workspace \{tag =
      NonNegative \{getNonNegative = 1\}, layout = -1, stack = Just (Stack \{focus
        = `\char`\\NUL', up = "", down = ""\})\},Workspace \{tag = NonNegative
      \{getNonNegative = 4\}, layout = -1, stack = Just (Stack \{focus = 'I', up =
        "", down = ""\})\}], floating = fromList []\} }
\end{sloppypar}
\medskip%
\noindent
(This counterexample uses Haskell's default \ttp{Show} instances, which uses
record syntax.)  Programmers may be familiar with having to debug
a ``wall of text'' as shown above.  What if instead a
formula like the following were returned, stating that for any well-typed
values $x_0$, $x_1$, $x_2$, and $x_3$, tested, a counterexample is found?

@@

For example, consider a simple calculator language containing constants,
addition, and division, together with an evaluator that checks if the divisor is
0 and returning \ttp{Nothing} in that case:
%
\begin{code}
data Exp = C Int
         | Add Exp Exp
         | Div Exp Exp

eval :: Exp -> Maybe Int
eval (C i) = Just i
eval (Add e0 e1) =
  liftM2 (+) (eval e0) (eval e1)
eval (Div e0 e1) =
  let e = eval e1 in
  if e == Just 0 then Nothing
    else liftM2 div (eval e0) e
\end{code}
%
\noindent
Now consider the property \ttp{prop\_div}, claiming that if \ttp{divSubTerms}
holds on an expression, then the evaluator returns \ttp{Just} a value:
%
\begin{code}
divSubTerms :: Exp -> Bool
divSubTerms (C _)         = True
divSubTerms (Div _ (C 0)) = False
divSubTerms (Add e0 e1)   =  divSubTerms e0
                          && divSubTerms e1
divSubTerms (Div e0 e1)   =  divSubTerms e0
                          && divSubTerms e1

>>>>

\exclusivelicense
\conferenceinfo{Haskell~'14}{September 6, 2014, Gothenburg, Sweden.}
\CopyrightYear{2014}
\copyrightdata{978-1-4503-3041-1/14/09}
\doi{2633357.2633365}

@@

\category{D.2.5}{Software Engineering}{Testing and Debugging}

@@

\keywords
property-based testing; test-case generalization; delta-debugging

@@

\begin{figure}[ht]
\begin{code}
\scriptsize
StackSet \{current = Screen \{workspace = Workspace
\{tag = NonNegative \{getNonNegative = 0\}, layout = -1, stack =
Just (Stack \{focus = `S', up ="", down = ""\})\}, screen = 1,
screenDetail = 1\}, visible = [Screen \{workspace = Workspace
\{tag = NonNegative \{getNonNegative = 2\}, layout = -1, stack =
Nothing\}, screen = 2, screenDetail = -1\},Screen \{workspace =
Workspace \{tag = NonNegative \{getNonNegative = 3\}, layout = -1,
stack = Nothing\}, screen = 0, screenDetail = -1\}], hidden =
[Workspace \{tag = NonNegative \{getNonNegative = 1\}, layout = -1,
stack = Just (Stack \{focus = `NUL', up = "", down = ""\})\},
Workspace \{tag = NonNegative \{getNonNegative = 4\}, layout = -1,
stack = Just (Stack \{focus = `I', up = "", down = ""\})\}],
floating = fromList []\}
\end{code}
  \caption{Counterexample returned by QuickCheck for a bug-injected version of \xmonad.}
  \label{fig:xmonadex}
\end{figure}

In addition, shrinking is one side of the coin when it comes to making
counterexamples more understandable: the other side is extrapolation from
individual counterexamples to a class of counterexamples.  This leap of
abstraction is often implicitly made by the programmer in determining the reason
why counterexamples fail the property.  For example, Figure~\ref{fig:xmonadex}
contains a relatively small counterexample returned when QuickChecking a
property in (a bug-injected version of) \xmonad, a popular X11 window manager
written in Haskell~\cite{xmonad}.  (This counterexample uses Haskell's default
\ttp{Show} instances, which uses record syntax.)  Programmers may be familiar
with having to debug a ``wall of text'' as shown above.  What if instead a
formula like the following were returned, stating that for any well-typed values
$x_0$, $x_1$, $x_2$, and $x_3$, tested, a counterexample is found?

%% %

%% \medskip%
%% \begin{sloppypar}
%% %\footnotesize
%% \noindent%
%% \ttp{StackSet \{current = Screen \{workspace = Workspace \{tag = NonNegative
%%       \{getNonNegative = 0\}, layout = -1, stack = Just (Stack \{focus = 'S', up =
%%         "", down = ""\})\}, screen = 1, screenDetail = 1\}, visible = [Screen
%%     \{workspace = Workspace \{tag = NonNegative \{getNonNegative = 2\}, layout = -1,
%%         stack = Nothing\}, screen = 2, screenDetail = -1\},Screen \{workspace =
%%       Workspace \{tag = NonNegative \{getNonNegative = 3\}, layout = -1, stack =
%%         Nothing\}, screen = 0, screenDetail = -1\}], hidden = [Workspace \{tag =
%%       NonNegative \{getNonNegative = 1\}, layout = -1, stack = Just (Stack \{focus
%%         = `\char`\\NUL', up = "", down = ""\})\},Workspace \{tag = NonNegative
%%       \{getNonNegative = 4\}, layout = -1, stack = Just (Stack \{focus = 'I', up =
%%         "", down = ""\})\}], floating = fromList []\} }
%% \end{sloppypar}
%% \medskip%
%% \noindent

@@

For example, consider a simple calculator language containing constants,
addition, and division, together with an evaluator that checks if the divisor is
0 and returning \ttp{Nothing} in that case:

\pagebreak

%
\begin{code}
data Exp = C Int
         | Add Exp Exp
         | Div Exp Exp
\end{code}
%
\begin{code}
eval :: Exp -> Maybe Int
eval (C i) = Just i
eval (Add e0 e1) =
  liftM2 (+) (eval e0) (eval e1)
eval (Div e0 e1) =
  let e = eval e1 in
  if e == Just 0 then Nothing
    else liftM2 div (eval e0) e
\end{code}
%
\noindent
Now consider the property \ttp{prop\_div}, claiming that if \ttp{divSubTerms}
holds on an expression, then the evaluator returns \ttp{Just} a value:
%
\begin{code}
divSubTerms :: Exp -> Bool
divSubTerms (C _)         = True
divSubTerms (Div _ (C 0)) = False
divSubTerms (Add e0 e1)   =  divSubTerms e0
                          && divSubTerms e1
divSubTerms (Div e0 e1)   =  divSubTerms e0
                          && divSubTerms e1
